---
title: "Blog Post 2"
author: "Ben LeStage"
date: "2024-01-31"
categories: [news, code, analysis]
---


```{r warning=FALSE, message=FALSE}
library(tidyverse)


tuesdata <- tidytuesdayR::tt_load('2024-01-09')

canada_births_1991_2022 <- tuesdata$canada_births_1991_2022
nhl_player_births <- tuesdata$nhl_player_births
nhl_rosters <- tuesdata$nhl_rosters
nhl_teams <- tuesdata$nhl_teams

```

### Introduction

For this post, I will be working with the Canadian Hockey Player Birth Months data from tidytuesday. 

https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-01-09/readme.md

Specifically, from this data I will be using the nhl_rosters data frame. This data set consists of 54,873 nhl players and the teams they played for. 

In this data frame I will be mainly looking at sweater_number, position_code, and shoots_catches. With this, I will be attempting to answer a couple questions and explore a few relationships in the data. My main question of interest consists of looking at the relationship between sweater_number and position_code. I am interested in this question because in my hockey career I have noticed a trend where Defensemen tend to wear single-didget numbers more often and I am curious as to see whether this trend that I have noticed is consistent in the NHL. Along with this, I will be looking at the relationship between shoots_catches and position_code to see if certain handedness is favored at specific positions in the NHL. I will also be looking at the relationship between shoots_catches and team_code to see if certain teams favor players of certain handedness.

### Primaary Visualizations

```{r}

ggplot(data = nhl_rosters, aes(x = sweater_number, fill = position_code)) +
  geom_bar() +
  coord_flip() +
  scale_fill_viridis_d()

```

This barplot demonstrates the trend in jersey numbers by position. As can be seen from the plot, the large majority of players throughout nhl history have worn jersey numbers that tend to be lower, regardless of position. This can probably be explained by the fact that in hockey, until relatively recently players only wore numbers between 1 and 30. Also demonstrated by the plot is that fact that numbers 1 and 30 tned to be worn by goalies rather than skaters. Again this speaks to the trend when players wore numbers between 1 and 30 with goalies typically wearing the jersey numbers 1 and 30 and the rest of the skaters wore numbers in between. Along with this, my hunch was correct that lower jersey numbers(specificially single didget numbers) tend to be worn by defencemen rather than forewards or goalies which is also demonstrated by the plot.


```{r}

ggplot(data = nhl_rosters, aes(x = shoots_catches, fill = position_code)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  scale_fill_viridis_d()

```

This barplot demonstrates some other interesting trends in NHL data. As can be seen from the plot, the NHL tends to be dominated by left-handed players which might seem surprising at first glance. However, in hockey players are coached to use their dominant hand on the top of the stick. For a right-handed person this would mean their right hand would be on the top of the stick which would make them a player that shoots left. This makes the barplot make more sense in that right handed people would be left-handed shooters and since there are more right handed people in the world, this trend makes sense. I was also surprised by the breakdown of handedness by position. Unsurprisingly, right handed shooters tend to play right wing more than they play left wing, this is also the case for left handed players playing left wing. Unsurprisingly, defensemen do not have as clear of a trend. This might make sense in the fact that defense is not broken up into left and right in the same way that the foreward position is. There are more left handed defenseman but agian this most likely just demonstrates the fact that there are more right hadned people in the world.


```{r}

ggplot(data = nhl_rosters, aes(x = team_code, fill = shoots_catches)) +
  geom_bar() +
  coord_flip() + 
  scale_fill_viridis_d()

```

This plot demonstrates a similar trend to the previous barplot. Again, this plot shows that the majority of players in the NHL shoot left, this can again be explained by the fact that the majority of people in the world are right handed. More interestingly however is the fact that it appears that some teams have a higher percentage of players that shoot right. For example, Toronto, Montreal, Detroit, Chicago, and Boston all have had more right handed players throughout their history than other teams. While this is an interesting discovery, this could also just be due to the difference in sample size. All of these teams are included in the original six NHL teams so they have been around the longest and therefore have had more players in their teams' histories.


### Conclusion and Wrap Up

Overall, I do not see many flaws with my approach to analyzing this data. Sample size is not an issue given that this data frame includes every player in NHL history. Along with this, I was not limited by any of the data points or in the organization of the data when making my visualizations. There were some missing data points for handedness, however I decided to include them as it gives an interesting perspective probably from a time when handedness of players was not recorded.


### Connections to class Ideas

Because all of my questions of interest were based around two categorical variables, I decieded to use stacked and side-by-side bar plots. I would have prefered to use side by side bar plots for each question of interest because it better allows someone to see the real count, however for the questions involving jersey numbers and teams there were too many categories on the x-axis making the plots far too messy.





















































